---
title: "Exploring NEON Plant and Soil Sampling"
author: "Angie Luis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE)

setwd("~/Documents/NEON data/NEONecology")
load("ReducedMergedPlantSoilData.RData")

library(tidyverse)
library(kableExtra)
library(neonDivData)
library(sf)
library(scales)
library(tigris)
library(data.table)


sitesDD <- neonDivData::neon_sites

```


I'm using the plant cover data from the neonDivData package and raw productivity and periodic soil sampling datasets downloaded with the neonUtilities package. I removed nlcdClass: cultivatedCrops, pastureHay, woodyWetlands, and emergentHerbaceousWetlands. I cut the data to 2019 and after. I only included June, July, and August in the plant sampling. For Soils / microbes, here I'm showing sampleTiming=="peakGreenness", but this still includes some sampling outside June-Aug, so we need to discuss. 

```{r summary, include=FALSE, echo=FALSE, warning=FALSE, results='hide', message=FALSE}


pc.sitehabitat <- reduced.plant.cover %>%
  group_by(siteID, nlcdClass) %>%
  summarise(n.cover.plots = length(unique(plotID)))

prod.sitehabitat <- productivity %>%
  group_by(siteID, nlcdClass) %>%
  summarise(n.prod.plots = length(unique(plotID)))

cover.productivity.sitehabitat <- full_join(pc.sitehabitat, prod.sitehabitat)

# filtering by peak greenness but discuss
soil.p.samples.sitehabitat <- soil.periodic.merge %>%
  filter(sampleTiming=="peakGreenness") %>%
  group_by(siteID, nlcdClass) %>%
  summarise(n.pchem = length(unique(sampleID[which(!is.na(CNratio))])), # there are more if don't need both C and N
            n.pmois = length(unique(sampleID[which(!is.na(soilMoisture))])),
            n.ppH = length(unique(sampleID[which(!is.na(soilInWaterpH))])),
            n.ITSseq = length(unique(sampleID[which(ITS.sequence=="Y" & ITS.qaqcStatus=="Pass")])))


soil.i.samples.sitehabitat <- soil.initial %>%
  group_by(siteID, nlcdClass) %>%
  summarise(n.iOlsenP = length(unique(plotID[which(!is.na(OlsenPExtractable))])), # there are more if don't need both C and N
            n.iMehlP = length(unique(plotID[which(!is.na(MehlichIIITotP))])),
            n.icaNh4d = length(unique(plotID[which(!is.na(caNh4d))])),
            n.itext = length(unique(plotID[which(!is.na(sandTotal))])))


plant.soil.sitehab.samples <- full_join(cover.productivity.sitehabitat, soil.p.samples.sitehabitat)
plant.soil.sitehab.samples <- full_join(plant.soil.sitehab.samples, soil.i.samples.sitehabitat)
plant.soil.sitehab.samples <- replace(plant.soil.sitehab.samples, is.na(plant.soil.sitehab.samples), 0)

sites <- sort(unique(as.character(plant.soil.sitehab.samples$siteID)))



dtc <- sitesDD[match(sites, sitesDD$siteID),]
dtc$color <-  hue_pal()(length(unique(dtc$siteID))) 
dt <- dtc[,c(2,7,6)]

## use shift_geometry to make map that has the territories
us_states <- states(cb = TRUE, resolution = "20m") %>%
  shift_geometry()

DT <- data.table(as.data.frame(dt))

DT_sf = st_as_sf(DT, coords = c("Longitude", "Latitude"), 
                 crs = 4326, agr = "constant")

new_site_coords <- DT_sf %>%
  shift_geometry()

```


```{r map, echo=FALSE, warning=FALSE, message=FALSE,  out.width='150%'}


ggplot() +
  geom_sf(data = new_site_coords, aes(colour = dtc$color[match(new_site_coords$siteID,dtc$siteID)])) +
  geom_sf(data = us_states, fill = NA, color = "black", size = 0.1) +
  geom_sf_label(data = new_site_coords, 
               aes(colour=dtc$color[match(new_site_coords$siteID,dtc$siteID)], label=siteID), label.size = 0.1, nudge_y = -110000, size=2.5) +
               #scale_fill_viridis_c() +
  theme_void(base_size = 16) +
  ggtitle("NEON terrestrial sites") +
    theme(plot.title = element_text(hjust = 0.5), legend.position = 'none') +
  scale_colour_identity()




```


## Summary of Sampling by Site-Habitat

Here is a summary of sampling of each site-habitat combo across datasets for plant cover, productivity, periodic and initial soils, ITS sequences. This is using the 'reduced.plant.cover' after removing a few - see below. 'n.cover.plots' is the number of unique plots sampled for plant cover (can be over 1 or multiple years and can have different number of subplots, etc). 'n.prod.plots' is the number of unique plots sampled for productivity (can be over 1 or multiple years and include multiple samples per season within and across subplots). "n.pchem", "n.pmois", and "n.ppH" are number of periodic soil samples (by 'sampleID') analyzed for the different soil attributes (chemistriy, moisture, pH). (Specifically, pchem here was CNratio - needed both data on C and N, and if only want one, or other chemistry, there may be more samples.) "n.ITSseq" is number of soil samples sequenced for microbes for ITS and passed qaqc. "iOlsenP", "iMehlP", "ichNh4d", and "itext" are number of plots sampled during the initial soil characterization (done once) for Olsen P, Mehlich P, chNh4d, and texture. 

```{r table, echo=FALSE, warning=FALSE, message=FALSE,  out.width='150%'}

plant.soil.sitehab.samples$state <- dtc$State[match(plant.soil.sitehab.samples$siteID,
                                                    dtc$siteID)]
plant.soil.sitehab.samples <- plant.soil.sitehab.samples[,c(1,13,2:12)]
psss.col <- dtc$color[match(plant.soil.sitehab.samples$siteID,dtc$siteID)]

#plant.soil.sitehab.samples <- plant.soil.sitehab.samples[order(c(plant.soil.sitehab.samples$siteID, plant.soil.sitehab.samples$nlcdClass)),]

plant.soil.sitehab.samples %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), fixed_thead = T, full_width = F) %>%
  kable_styling(font_size=13, position = "left", full_width = FALSE) %>%
  column_spec(1, bold = TRUE, border_right = TRUE, color = psss.col) %>% #, background = "grey85") %>%
  column_spec(4, color = ifelse(plant.soil.sitehab.samples$n.cover.plots==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.cover.plots==0, TRUE, FALSE)) %>%
  column_spec(5, color = ifelse(plant.soil.sitehab.samples$n.prod.plots==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.prod.plots==0, TRUE, FALSE)) %>%
  column_spec(6, color = ifelse(plant.soil.sitehab.samples$n.pchem==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.pchem==0, TRUE, FALSE)) %>%
  column_spec(7, color = ifelse(plant.soil.sitehab.samples$n.pmois==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.pmois==0, TRUE, FALSE)) %>%
  column_spec(8, color = ifelse(plant.soil.sitehab.samples$n.ppH==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.ppH==0, TRUE, FALSE)) %>%
  column_spec(9, color = ifelse(plant.soil.sitehab.samples$n.ITSseq==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.ITSseq==0, TRUE, FALSE)) %>%
  column_spec(10, color = ifelse(plant.soil.sitehab.samples$n.iOlsenP==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.iOlsenP==0, TRUE, FALSE)) %>%
  column_spec(11, color = ifelse(plant.soil.sitehab.samples$n.iMehlP==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.iMehlP==0, TRUE, FALSE)) %>%
  column_spec(12, color = ifelse(plant.soil.sitehab.samples$n.icaNh4d==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.icaNh4d==0, TRUE, FALSE)) %>%
  column_spec(13, color = ifelse(plant.soil.sitehab.samples$n.itext==0, "firebrick2", "black"),
              bold = ifelse(plant.soil.sitehab.samples$n.itext==0, TRUE, FALSE)) 

    
    # column_spec(1, bold = TRUE, border_right = TRUE, color = "black", background = "lightgrey")
   

```

See "PlantSoilSiteHabitatYearSampling.xlsx" in UMT Box folder for same thing but also separated out by year.

## For each habitat, how many sites have ITS microbial sequence data and other variables?
This many sites have at least one year where there is data for plant cover, productivity, ITS sequences, initial soil sampling for Mehlich P, and texture. (I removed periodic soils for this - but there will be some periodic soil data for all these as well since happens at same time as microbial sampling, just maybe not every variable e.g, might have moisture but not chemistry)
```{r sh, echo=FALSE, warning=FALSE, message=FALSE}

 
replace(plant.soil.sitehab.samples, plant.soil.sitehab.samples==0, NA) %>%
  select(c(siteID, nlcdClass, n.cover.plots, n.prod.plots, n.ITSseq, n.iMehlP, n.itext)) %>%
  drop_na() %>%
  group_by(nlcdClass) %>%
  summarise(n.sites = n()) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), fixed_thead = T, full_width = F)


```


<br/>
<br/>

## Exploring sampling of Productivity

#### how many productivity samples per plot per year?
```{r ppy, echo=FALSE, warning=FALSE, message=FALSE}

# how many productivity samples per plot per year?
prod.plotyear.samples <- productivity %>%
  group_by(siteID, plotID, nlcdClass, year(collectDate)) %>%
  summarise(n.subplots = length(unique(subplotID)),
    n.samples = length(unique(sampleID))) 
summary(factor(prod.plotyear.samples$n.samples))            
#   1    2    3    4    5    6    7    8    9 
# 743 1047   47   93    1   21    9    1   10 
# usually 2 samples per plot per year, but again a range
#hist(summary(prod.plotyear.samples$n.samples))

```
[Top row is the number of samples, and bottom row is how many plot-years had that many of that number of samples. E.g., 1047 plot-years had 2 samples, and 93 had 4 samples. ]
Usually 2 samples per plot per year, but a range of up to 9

#### how many productivity subplots per plot per year?
```{r psy, echo=FALSE, warning=FALSE, message=FALSE}

summary(factor(prod.plotyear.samples$n.subplots))            
#   1   2   3 
# 995 976   1 
# sometimes 1, sometimes 2 subplots per plot per year
#hist(summary(prod.plotyear.samples$n.subplots))

```
sometimes 1, sometimes 2 subplots per plot per year (once 3)


#### how many productivity samples per subplot per year?
```{r ppsy, echo=FALSE, warning=FALSE, message=FALSE}

# number of samples per subplot per year
prod.subplotyear.samples <- productivity %>%
  group_by(siteID, plotID, subplotID, nlcdClass, year=year(collectDate)) %>%
  summarise(n.samples = length(unique(sampleID))) 
summary(factor(prod.subplotyear.samples$n.samples))            
#    1    2    3    4    5    6    7    9 
# 2674  102   47   87    1   21    9    9 
#hist(summary(prod.subplotyear.samples$n.samples))            

```
Mostly 1 but up to 9.

####  do the ones that have more than 1 sample per subplot per year have exclosure treatments?

```{r rpsx1, echo=FALSE, warning=FALSE, message=FALSE}

ps <- unique(prod.subplotyear.samples$plotID[which(prod.subplotyear.samples$n.samples>1)])

y <- productivity %>%
  filter(plotID %in% ps) %>%
  group_by(siteID, plotID, year=year(collectDate)) %>%
  summarise(any.Y = factor(ifelse(length(which(exclosure=="Y"))>0, "Y","N")))

summary(y$any.Y)

# N   Y 
# 52 251 

```
Not all of them, but most (5/6).

### More about the exclosure treatments

####  Which sites have exclosures?
```{r psx2, echo=FALSE, warning=FALSE, message=FALSE}

as.character(unique(productivity$siteID[productivity$exclosure=="Y"]))
    
  

```
12 out of 41 sites have at least one exclosure over the time period.

#### What habitat types?
```{r psx3, echo=FALSE, warning=FALSE, message=FALSE}

dat <- productivity %>%
  filter(exclosure=="Y") %>%
  group_by(siteID, nlcdClass, plotID) %>%
  summarise(n.clipIDs= length(unique(clipID)))
unique(dat$nlcdClass)
  

```
From 1 to 9 clipIDs from within exlcosures per site-habitat.

#### What's the spatial scale?

Here's an example:

```{r psxy, echo=FALSE, warning=FALSE, message=FALSE}

# Is exclosure defined at the plot level or subplot?

dat <- productivity %>%
  group_by(siteID, plotID, year=year(collectDate)) %>%
  summarise(num.Y = length(which(exclosure=="Y")),
            num.N = length(which(exclosure=="N")))
# summary(dat)
# dat[which(dat$num.Y>0 & dat$num.N>0),]
# plots for which some subplot some year have more than 1 sample

p <- "CLBJ_004" # "CLBJ_010"   
y <- 2021       # 2019        

productivity %>%
  mutate(year = year(collectDate)) %>%
  select(plotID, subplotID, year, collectDate, exclosure) %>%
  filter(plotID==p, year==y) %>%
  distinct()



dat <- productivity %>%
    mutate(year = year(collectDate)) %>%
    filter(plotID==p, year==y) %>%
    group_by(plotID, subplotID, year, collectDate, clipID) %>%
    summarise(excl = factor(ifelse(length(unique(exclosure))==1, "1 result", "more")))

  
```
Same plotID and subplotID with some dates having both exclosure and no exclosure. There were more than 1 clip samples taken per subplot and exclosure must be at a finer spatial scale than subplot. The same clipID or sampleID don't have both a Y or N for exclosure. 

#### For every Y exclosure clipID is there a corresponding N for the subplot for the date?
(or within a date range?)
```{r psx4, echo=FALSE, warning=FALSE, message=FALSE}
excl.plots <- unique(productivity$plotID[which(productivity$exclosure=="Y")])

dat <- productivity %>%
  filter(plotID %in% excl.plots) %>%
  group_by(plotID, subplotID, collectDate) %>%
  summarise(num.Y = length(which(exclosure =="Y")), # this won't give unique samples (because more than 1 row per sample)
            num.N = length(which(exclosure=="N")))
dat <- dat %>%
  filter(num.Y !=0)

dat[dat$num.N==0,]


dat2 <- productivity %>%
  filter(plotID %in% excl.plots) %>%
  group_by(plotID, year = year(collectDate)) %>%
  summarise(num.Y = length(which(exclosure =="Y")),
            num.N = length(which(exclosure=="N")))
length(which(dat2$num.N==0))

```
Most of the time, for every Y there is a (At least 1) corresponding N on the same subplot on the same date. But not for 3. [but within the same year]

**Overall, for Productivity, I will average total biomass per clipID per plot per year. If exclosures present, I will assume that this average will be a mix of exclosure and no exclosure. For clipIDs with exclosures, it looks like they only quantified "All herbaceous plants", but for others they separated out into groups. I will assume that a sum over all categories is equivalent to "all", for a general productivity measure.**

May later want to quantify specific functional groups, but will have to think about that later.

<br/>
<br/>

## Plant Cover 

### How many subplots per plot per year?
These are all per year, and based on the subplotID name, which looks like for example, "41_1_4" for 1m^2^ subplot and "41_10_4" for the 10m^2^ subplot.


#### how many 100m^2^ subplots per plot?
```{r cov3, echo=FALSE, warning=FALSE, message=FALSE}
plant.100mplotyears <- plant.cover[grep("_100",plant.cover$subplotID),] %>%
  group_by(siteID, plotID, year=year(observation_datetime)) %>%
  summarise(num.100m2.plots=length(unique(subplotID)))
summary(factor(plant.100mplotyears$num.100m2.plots))
# 1    2    3    4 
# 4    6   68 1399 

# example 
# p="BARR_001"
# y=2022
# 
# plant.cover %>%
#   filter(plotID==p, year(observation_datetime)==y) %>%
  

```
There should be 4 and were most of the time, but sometimes less.

#### how many 10m^2^ subplots per plot?
```{r cov1, echo=FALSE, warning=FALSE, message=FALSE}


# how many 10m2 plots per plot?
plant.10mplotyears <- plant.cover[grep("_10_",plant.cover$subplotID),] %>%
  group_by(siteID, plotID, year=year(observation_datetime)) %>%
  summarise(num.10m2.plots=length(unique(subplotID)))
summary(factor(plant.10mplotyears$num.10m2.plots))
# 1    2    3    4    5    6 
# 2    2   15   34   89 1335 
```
mostly 6 subplots but a few with less


I think it's ok that there are sometimes fewer 10m^2^ or 100m^2^ subplots per plot than expected (6 or 4). Because if there was no species that hadn't been identified previously in the smaller subplots, then there would be no line of data. It's possible that a line of comments was removed when cleaning from raw data to divData.

However, 1m^2^ not ok. Those should all be quantified for % cover so shouldn't be any missing.


#### how many 1m^2^ subplots per plot?
```{r cov2, echo=FALSE, warning=FALSE, message=FALSE}

plant.1mplotyears <- plant.cover[grep("_1_",plant.cover$subplotID),] %>%
  group_by(siteID, plotID, year=year(observation_datetime)) %>%
  summarise(num.1m2.plots=length(unique(subplotID)))
summary(factor(plant.1mplotyears$num.1m2.plots))
# 1    2    3    4    5    6 
# 2    9   24   53  114 2768 

# example
p="ABBY_019"
y=2020

dat <- plant.cover %>%
  filter(plotID==p, year(observation_datetime)==y)

# as.data.frame(dat)
# 
# unique(dat$subplotID)

```
Looking at one example, plot="ABBY_019", year=2020: there were 4 1m^2^ subplots and no larger (10 or 100) subplots. That should be eliminated. **I will eliminate all the plot-years that didn't have 6 1m^2^-subplots.** (3% of cover data.) Now called reduced.plant.cover.



#### how many of any size subplots per plot?
```{r cov4, echo=FALSE, warning=FALSE, message=FALSE}

plant.anysubplotyears <- reduced.plant.cover %>%
  group_by(siteID, plotID, year=year(observation_datetime)) %>%
  summarise(num.subplots=length(unique(subplotID)))
summary(factor(plant.anysubplotyears$num.subplots))
#    6   11   12   13   14   15   16 
# 1410    2    6    4   19   84 1244 



```

Since, here I already removed the plot-years that didn’t have 6 1m^2^-subplots, the ones here that have 6 are ones where there aren't any larger (10 and 100sqm) plots in the data. (Could be because they didn't ID any additional species at the larger plots that weren't in the 1m^2^ plot.)

#### After cutting to June/July/August is there more than one bout per plot per year?
What they called "boutNumber"

```{r bout, echo=FALSE, warning=FALSE, message=FALSE}
dat <- reduced.plant.cover %>%
  group_by(siteID, plotID, year=year(observation_datetime)) %>%
  summarise(n.bouts = length(unique(boutNumber)))
summary(factor(dat$n.bouts))
```
No. There is only 1 bout per plot per year now.

<br/>


**The unit I am calculating is species richness per plot per year. So total number of unique species identified at any of the subplots within a plot per year.**

We can do rarefaction per site/habitat or something else to get a measure to use.


<br/>
<br/>

## Periodic Soil Characteristics & Microbe Sequencing 

### 

The Soil Chemistry dataframe sometimes had multiple rows per sampleID. E.g., there were sometimes 2 different testMethods to get C vs N, each with its own row. So there are NAs for nitrogen but a value of carbon or vice versa on different rows. Sometimes 2 different values for the same variable. For  d15N, organicd13C, nitrogenPercent, organicCPercent, and CNratio, I took the average (removing NAs) so that there was one value per sampleID to match up with the rest of the data. I merged the data into one dataframe called soil.periodic.merge with one row per sampleID, with information on plot, date, sampleTiming, horizon, depth, chemistry, moisture, pH, and whether it was sequenced for ITS and passed qaqc.


<br/>
<br/>

## Initial Soil Characterization

Happened once. 
Here horizonID is unique identifier that I merged datasets with. Each plot has one pit with multiple horizons. I merged info on plots and horizons with biogeochem, bulk density, and particle size. The bulk density sometimes had more than 1 test run (clod vs compliant cavity) on the same sample (horizonID), so I again took the average per horizonID (removing NAs).

I tried to align horizons between the soil periodic sampling (and microbe sampling) and initial soil characterizations. The plots aren't always present in both datasets so used any plots in the same site-habitat from the initial soils. More than half of the microbe sample top and bottom depths were not fully within an initial soil horizon. But if just looking for any overlap in the 2 horizons, then only 45 out of 6085 that didn't line up at all (could choose the closest?).



